Interpretable Machine Learning

### 2018-10-23 poster/poster_2018.Rnw
adds more images

<<rent>>=
knitr::include_graphics("figure/rent.png")
@

@@

\block{Feature effects}{
The feature effect describe how changing a features changes the prediction of all data points on average. \cite{friedman2001greedy}
<<ice, echo=FALSE, fig.width=12, fig.height=6>>=
eff = FeatureEffect$new(pred, "size", method = "pdp")
eff.res = eff$results
ggplot(eff.res)  + geom_line(aes(x=size, y = .y.hat), size = 3) +
  scale_x_continuous("Size of living area") + 
  scale_y_continuous("Predicted rent") +
  theme(aspect.ratio=2/4, text = element_text(size=35),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
@
For example we can vary the size of the living area and observe how the prediction of the model changes.

@@

ggplot(impres) + geom_col(aes(y = importance, x = feature)) + coord_flip() +
  theme(aspect.ratio=2/4, text = element_text(size=30),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_x_discrete("") +
  scale_y_continuous("Feature Importance")
@
In the rent example the most important feature was the size of the living area. 
}
\block{Individual predictions}{

@@

We have a couple of options to explain why a particular prediction was made by a machine learning model.
Suppose we want to explain the prediction of the following appartment. 
<<rent2>>=
knitr::include_graphics("figure/rent.png")
@

<<counterfact, echo=FALSE, include = FALSE>>=
appartment.mod = appartment
appartment.mod$cats = factor("No", levels = c("No", "Yes"))
pred.new = round(pred$predict(appartment.mod[,1:3, drop=FALSE]), 0)
@
\vspace{1cm}
\section*{Counterfactual explanations}
A counterfactual explanation of a prediction describes the smallest change to the feature values (inputs) that changes the prediction to a predefined output.
For example: How does the inputs have to change so that the prediction changes below 900 Euros?
Answer (one of many possible ones): If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.

\section*{Shapley Values}
From a game theoretical perspective, the feature values can be seen as players in a game that cooperate to achieve the prediction as a payout.
The Shapley Values method \cite{vstrumbelj2014explaining} fairly attributes the difference between the prediction and the average prediction among the feature values.

<<shapley, echo=FALSE, fig.width=12, fig.height=6>>=
shap = Shapley$new(pred)
shap$explain(appartment)
plot(shap) +
  theme(aspect.ratio=2/4, text = element_text(size=30),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_y_continuous("Contribution to prediction minus avg. prediction") +
  scale_x_discrete("")

@@

\block{Interpretable Models}{
Intrinsically interpretable models learn simple relationships between input and output,
but often predict the data less well than flexible, yet less interpretable, models.

\section*{Linear Regression Models}
%The linear regression model models the outcome y as a weighted sum of the input features.

@@

%The goal is to find the best possible weights $\beta$.
<<lm>>=
knitr::include_graphics("figure/lm.png")
@

In the rent example, the following model is learned.
<<train-model-lm, echo=FALSE, include=FALSE>>=
mod.lm = lm(rent ~ ., data = df)
cf = round(coef(mod.lm))
knitr::kable(cf, format = 'latex')
@

$$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  + \Sexpr{cf["locationGood"]} \cdot I_{\text{good location}} +  \Sexpr{cf["catsYes"]} \cdot I_{\text{cats allowed}}$$

An increase of the living area increases the predicted rent by \Sexpr{coef(mod.lm)["size"]}.
Changing the location from "Bad" to the "Good" increases the predicted rent by \Sexpr{coef(mod.lm)["locationGood"]}.
%Linear models have many extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity.

\section*{Decision Trees}
Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.

<<train-model-tree, echo=FALSE, out.width="0.28\\textwidth">>=
library(partykit)
mod.tree = ctree(rent ~ ., data = df, control = ctree_control(maxdepth = 2))
#plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
knitr::include_graphics("figure/tree.png")
@
For example, the tree predicts a rent of 1908 Euros for an appartment bigger than 77 $m^2$ in a good location.

\section*{Decision Rules}
A decision rule is a simple IF-THEN statement consisting of a condition and a prediction.
One or many rules can be used to make predictions.

@@

One decision rule learned for the rent prediction is:

IF living area $\in 90 \text{ and } 110 m^2$  AND location $=$ Good THEN the rent is between 1540 and 1890 EUR

\section*{Global Surrogate Models}
Intrinsically interpretable models can also be used as to increase interpretability of a black box model by approximating its predictions, acting as a surrogate model which can be interpreted. 
Example: Neural network as black box model, decision tree as surrogate model.

\section*{Local Surrogate Models}
The idea behind local surrogate models \cite{ribeiro2016should} is to replace the complex model with an individual, locally weighted interpretable model (e.g. decision tree) for the data point for which we want to explain the prediction.

<<lime, echo=FALSE>>=
# locmod = LocalModel$new(pred, k = 2)
# locmod$explain(df[1,1:3])
# plot(locmod)
knitr::include_graphics("figure/local-surrogate.png")
@
}

>>>>

<<rent>>=
knitr::include_graphics("figure/rent-ml.png")
@

@@

\block{Feature effects}{
The feature effect describe how changing a features changes the prediction of all data points on average. \cite{friedman2001greedy}
<<ice, echo=FALSE, fig.width=12, fig.height=6>>=
eff = FeatureEffect$new(pred, "size", method = "pdp")
eff.res = eff$results
ggplot(eff.res)  + geom_line(aes(x=size, y = .y.hat), size = 3) +
  scale_x_continuous("Size of living area") + 
  scale_y_continuous("Predicted rent") +
  theme(aspect.ratio=2/4, text = element_text(size=35),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank())
@
\textit{For example we can vary the size of the living area and observe how the prediction of the model changes.}

@@

ggplot(impres) + geom_col(aes(y = importance, x = feature)) + coord_flip() +
  theme(aspect.ratio=2/4, text = element_text(size=30),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_x_discrete("") +
  scale_y_continuous("Feature Importance")
@
\textit{In the rent example the most important feature was the size of the living area. }
}
\block{Individual predictions}{

@@

We have a couple of options to explain why a particular prediction was made by a machine learning model.
Suppose we want to explain the prediction of the following appartment. 
<<rent2, out.width="0.2\\textwidth">>=
knitr::include_graphics("figure/rent.png")
@

<<counterfact, echo=FALSE, include = FALSE>>=
appartment.mod = appartment
appartment.mod$cats = factor("No", levels = c("No", "Yes"))
pred.new = round(pred$predict(appartment.mod[,1:3, drop=FALSE]), 0)
@
\\ \\
\textbf{Counterfactual explanations} \\
A counterfactual explanation of a prediction describes the smallest changes to the feature values (inputs) that change the prediction to a predefined output.
For example: How does the inputs have to change so that the prediction changes below 900 Euros?
Answer (one of many possible ones): \\
\textit{If cats were allowed, the predicted rent would be \Sexpr{pred.new} Euros.}
\\ \\
\textbf{Shapley Values} \\
From a game theoretical perspective, the feature values can be seen as players in a game that cooperate to achieve the prediction as a payout.
The Shapley Values method \cite{vstrumbelj2014explaining} fairly attributes the difference between the prediction and the average prediction among the feature values.

<<shapley, echo=FALSE, fig.width=12, fig.height=6>>=
shap = Shapley$new(pred)
shap$explain(appartment)
plot(shap) +
  theme(aspect.ratio=2/4, text = element_text(size=30),
    panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_y_continuous("Contribution to difference of \nprediction minus avg. prediction") +
  scale_x_discrete("")

@@

\block{Interpretable Models}{
Intrinsically interpretable models learn simple relationships between input and output,
but often predict the data less well than flexible, yet less interpretable, models.
\\ \\
\textbf{Linear Regression Models}\\
%The linear regression model models the outcome y as a weighted sum of the input features.

@@

%The goal is to find the best possible weights $\beta$.
<<lm>>=
knitr::include_graphics("figure/lm.png")
@
\textit{In the rent example, the following weights were learned:}
<<train-model-lm, echo=FALSE, include=FALSE>>=
mod.lm = lm(rent ~ ., data = df)
cf = round(coef(mod.lm))
knitr::kable(cf, format = 'latex')
@

$\text{rent} = \Sexpr{cf["(Intercept)"]} + \Sexpr{cf["size"]} \cdot \text{size}  + \Sexpr{cf["locationGood"]} \cdot I_{\text{good location}} +  \Sexpr{cf["catsYes"]} \cdot I_{\text{cats allowed}}$
\textit{The predicted rent increases by \Sexpr{cf["size"]} Euros per each $m^2$.}
%Changing the location from "Bad" to the "Good" increases the predicted rent by \Sexpr{coef(mod.lm)["locationGood"]}.
%Linear models have many extensions, like Generalized Linear Models (GLMs) or Generalized Additive Models (GAMs) that allow more flexiblity.
\\ \\
\textbf{Decision Trees}\\
Decision trees partition the data into smaller subsets, based on 'decisions' made on the input features.

<<train-model-tree, echo=FALSE, out.width="0.28\\textwidth">>=
library(partykit)
mod.tree = ctree(rent ~ ., data = df, control = ctree_control(maxdepth = 2))
#plot(mod.tree, inner_panel = node_inner(mod.tree, pval = FALSE), type='simple')
knitr::include_graphics("figure/tree.png")
@
\textit{The tree predicts a rent of 1908 Euros for an appartment bigger than 77 $m^2$ in a good location.}
\\ \\
\textbf{Decision Rules} \\
Decision rules are IF-THEN statements consisting of a condition and a prediction.
One or many rules can be used to make predictions. For example:

@@

\textit{
IF $90m^2\leq \text{size} < 110m^2$  AND location $=$ Good THEN the rent is between 1540 and 1890 EUR}

}
\block{Surrogate Models}{
\textbf{Global Surrogate Models} \\
Intrinsically interpretable models (e.g. tree)  can also be used as to increase interpretability of a black box model (e.g. neural network) by approximating its predictions, acting as a surrogate model which can be interpreted. 
<<surrogate>>=
knitr::include_graphics("figure/global-surrogate.png")
@
\\ \\
\textbf{Local Surrogate Models}\cite{ribeiro2016should} \\
Replace a  complex model with an individual, locally weighted interpretable model (e.g. decision tree) for the data point for which we want to explain the prediction.

<<lime, echo=FALSE, out.width="0.2\\textwidth">>=
# locmod = LocalModel$new(pred, k = 2)
# locmod$explain(df[1,1:3])
# plot(locmod)
knitr::include_graphics("figure/local-surrogate.png")
@
}
